### 2021-01-05 ch03_esmvaltool/esmvaltool.tex
Added contributions to ESMValCore to chapter 03

In order to answer the research questions of this thesis posed in
\cref{sec:01:research_questions}, a reliable and efficient tool to read,
process and evaluate climate model output and observational data is necessary.
A valuable software that fits these criteria is the \ac{ESMValTool}. The
\ac{ESMValTool} is an open-source community diagnostics and performance metrics
tool for the routine evaluation of \ac{ESM} output, which notably facilitates
the analysis of \ac{CMIP} models (\url{https://www.esmvaltool.org}). For this
reason, all analyses presented in
\cref{ch:04:assessment_climate_metrics,ch:05:paper_ecs,ch:06:paper_gpp} of this
thesis were implemented into the \ac{ESMValTool}. Apart from that, further
substantial contributions to the code base of the \ac{ESMValTool} have been
made that improve the routine evaluation of climate models, which is beneficial
for the entire scientific community. Moreover, this lead to co-authorship in
the scientific documentation of the \ac{ESMValTool}, which is published in four
peer-reviewed studies \autocite{Eyring2020, Lauer2020, Righi2020, Weigel2020}.
After a brief overview over the \ac{ESMValTool}'s structure, this chapter
presents all these contributions that are not documented in other chapters of
this thesis.

@@

- Changes in \ac{ESMValCore} + table

>>>>

In order to answer the research questions of this thesis posed in
\cref{sec:01:research_questions}, a reliable and efficient tool to read,
process and evaluate climate model output and observational data is necessary.
A valuable software that fits these criteria is the \ac{ESMValTool}. The
\ac{ESMValTool} is an open-source community diagnostics and performance metrics
tool for the routine evaluation of \ac{ESM} output, which notably facilitates
the analysis of \ac{CMIP} models (\url{https://www.esmvaltool.org}). For this
reason, all analyses presented in
\cref{ch:04:assessment_climate_metrics,ch:05:paper_ecs,ch:06:paper_gpp} of this
thesis were implemented into the \ac{ESMValTool}. Apart from that, further
substantial changes and additions to the code base of the \ac{ESMValTool} have
been contributed that improve the routine evaluation of climate models, which
is beneficial for the entire scientific community. This lead to co-authorship
in the scientific documentation of the \ac{ESMValTool}, which is published in
four peer-reviewed studies \autocite{Eyring2020, Lauer2020, Righi2020,
  Weigel2020}. After a brief overview over the \ac{ESMValTool}'s structure,
this chapter presents all these contributions that are not documented in other
chapters of this thesis.

@@

As of December 2020, $40408$ lines of code have been added and $21050$ lines of
code have been removed from to the GitHub repository of \ac{ESMValCore} by the
author of this thesis. Apart from general improvements of the code base, these
changes and additions mainly include new preprocessor functions that can be
applied to the input data and derivation scripts for new non-\ac{CMOR}
variables. \Cref{tab:03:changes_to_esmvalcore} shows a summary of these main
contributions, which are partly published in the scientific documentation of
\ac{ESMValCore} \autocite{Righi2020}.

\begin{table}[t]
  \centering
  \begin{tabulary}{\columnwidth}{L L L}
    \toprule
    Type & Name & Description \\
    \midrule
    Preprocessor & \emph{amplitude} & Amplitude of periodic phenomena (\eg{}
    diurnal or seasonal cycles) \\
    \cmidrule(l){2-3}
    & \emph{land/sea fraction weighting} & Weighting of fields based on
    the land or sea fraction of the respective grid cells \\
    \cmidrule(l){2-3}
    & \emph{trend} & Linear trend (slope of ordinary least squares regression)
    \\
    \midrule
    Derived variable & \emph{asr} & Absorbed shortwave radiation \\
    \cmidrule(l){2-3}
    & \emph{co2s} & Atmospheric \ac{CO2} concentration at
    surface \\
    \cmidrule(l){2-3}
    & \emph{et} & Evapotranspiration \\
    \cmidrule(l){2-3}
    & \emph{rlntcs} & Clear-sky net \acf{TOA} longwave radiation \\
    \cmidrule(l){2-3}
    & \emph{rsntcs} & Clear-sky net \ac{TOA} shortwave radiation \\
    \cmidrule(l){2-3}
    & \emph{uajet} & Position of austral jet stream \\
    \bottomrule
  \end{tabulary}
  \caption{Summary of changes and features contributed to the
    \acf{ESMValCore}.}
  \label{tab:03:changes_to_esmvalcore}
\end{table}

In total, three new preprocessor functions have been added: \emph{amplitude},
\emph{land/sea fraction weighting} and \emph{trend}. The \emph{amplitude}
preprocessor calculates the peak-to-peak amplitude of periodic phenomena along
an arbitrary coordinate. Usually, this coordinate is a temporal dimension. A
common application of this preprocessor is the calculation of a variable's
diurnal or seasonal cycle amplitude. The \emph{land/sea fraction weighting}
preprocessor weights fields with the land or sea fraction of the respective
grid cells. For example, this is necessary for the spatial integration of
flux-related variables which are reported in units of \enquote{per square meter
  of land/sea} and not in \enquote{per square meter of grid cell}. After the
weighting, the grid cell areas can be used to integrate the flux-related
variable over a desired region. The \emph{trend} preprocessor calculates the
linear trend of a variable along an arbitrary coordinate. The linear trend is
defined as the slope of an ordinary least squares linear regression of the
variable against the selected coordinate. For example, this can be used to
calculate the temporal trend of the \ac{GSAT} over the \nth{20} century.

\begin{figure}[t]
  \centering
  \begin{subfigure}[t]{\SubfigureWidth{}}
    \includegraphics[width=\columnwidth]{ch03_esmvaltool/figs/co2s_map.pdf}
    \caption{}
    \label{fig:03:co2s:a}
  \end{subfigure}
  ~
  \begin{subfigure}[t]{\SubfigureWidth{}}
    \includegraphics[width=\columnwidth]{
      ch03_esmvaltool/figs/co2s_mlo_cycle.pdf}
    \caption{}
    \label{fig:03:co2s:b}
  \end{subfigure}
  \caption{(a) \acs{CMIP}6 \acf{MMM} of the global atmospheric \acs{CO2}
    concentration at surface (\emph{co2s}) averaged over the months \acf{JJA}
    of the year 2014 in the emission-driven historical simulation. The
    \acs{CMIP}6 \ac{MMM} includes the four climate models that provide all
    data needed for the calculation of \emph{co2s} (see legend in (b)). (b)
    Monthly-mean \emph{co2s} at Mauna Loa, Hawaii ($19.5 \unit{\degree N}$,
    $155.6 \unit{\degree W}$; elevation: $3397 \unit{m}$) from 1958 to 2014.
    The thick black line shows observations from the Mauna Loa Observatory
    \autocite{Keeling2005}; the remaining colored lines show emission-driven
    historical simulations from individual \acs{CMIP}6 models.}
  \label{fig:03:co2s}
\end{figure}

Apart from new preprocessor functions, six derivation scripts for the
non-\ac{CMOR} variables \emph{asr}, \emph{co2s}, \emph{et}, \emph{rlntcs},
\emph{rsntcs} and \emph{uajet} have been added. The absorbed shortwave
radiation (\emph{asr}) is defined as the difference in the incoming \ac{TOA}
shortwave radiation and the outgoing \ac{TOA} shortwave radiation. The
atmospheric \ac{CO2} concentration at surface (\emph{co2s}) can be calculated
from the pressure level-dependent atmospheric \ac{CO2} concentration and the
surface air pressure using interpolation. An example illustrating \emph{co2s}
is given in \cref{fig:03:co2s}. \Cref{fig:03:co2s:a} shows the \acs{CMIP}6
\ac{MMM} of the global \emph{co2s} for the months \acf{JJA} of the year 2014 in
the emission-driven historical simulation. As expected, the map shows high
atmospheric \ac{CO2} concentrations over large metropolitan areas with high
\ac{CO2} emissions (\eg{} over North America, the Arabian Peninsula and East
Asia). Moreover, regions with high photosynthetic activity (\eg{} the boreal
forests in the Northern hemisphere summer and the tropical rainforests in South
America) exhibit smaller \ac{CO2} concentrations. \Cref{fig:03:co2s:b} shows
the monthly-mean \emph{co2s} at Mauna Loa, Hawaii from 1958 to 2014 for
observations performed at the Mauna Loa Observatory \autocite{Keeling2005}
(thick black line) and emission-driven historical simulations from four
\acs{CMIP}6 models (remaining colored lines). Apart from the model CNRM-ESM2-1,
the simulated atmospheric \ac{CO2} concentrations show the expected seasonal
cycle present in the Keeling Curve (see \cref{fig:02:keeling_curve}). All
climate models correctly simulate the increase in \emph{co2s} over the years.
However, their are some differences in the absolute values with the model
BCC-CSM2-MR showing the largest deviations.

The evapotranspiration (\emph{et}) defined as the sum of evaporation and plant
transpiration can be calculated from the surface latent heat flux and the
latent heat vaporization constant. The longwave/shortwave clear-sky net
\acf{TOA} radiations (\emph{rlntcs}/\emph{rsntcs}) is defined as the difference
between the longwave/shortwave incoming \ac{TOA} radiation assuming clear-sky
and the longwave/shortwave outgoing \ac{TOA} radiation assuming clear-sky.
Finally, the position of the austral jet stream (\emph{uajet}) is given by the
latitude with maximum zonal-mean eastward wind speed at $850 \unit{hPa}$ on the
Southern hemisphere between $80 \unit{\degree S}$ and $30 \unit{\degree S}$.
